---
title: "G3 example - DEMO (R Version)"
output: html_document
---

# Particle Size Data Processing

In this challenge, we process and summarize particle size data using the `tidyverse`. This workflow is designed to handle multiple raw text files, perform geometric calculations, and bin data for downstream analysis.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(writexl)
```

## Setup Paths and Metadata

Specify the path and file name of the metadata file. This file should contain a list of filenames and their corresponding batches.

```{r}
# Path updated for relative access within the repository
filenames_path <- "./"
filenames_name <- "Raw data & meta data.xlsx"

# Note: This step assumes the metadata file exists.
# We wrap it in a conditional or provide a template for students.
if (file.exists(paste0(filenames_path, filenames_name))) {
  df_names <- read_excel(paste0(filenames_path, filenames_name), sheet = 'Data for Python DEMO')
} else {
  message("Metadata file not found. Please provide 'Raw data & meta data.xlsx'.")
  # Placeholder for demonstration
  df_names <- tibble(Filename = character(), Batch = character())
}
```

## Create one stacked file combining all individual files

In R, we can use `purrr::map_df` to read multiple files, apply sampling, and stack them efficiently.

```{r}
# This chunk demonstrates how to read and sample 5000 rows from each file
read_and_sample <- function(row) {
  if (!file.exists(row$Filename)) return(NULL)

  read_tsv(row$Filename, locale = locale(encoding = "latin1")) %>%
    slice_sample(n = 5000) %>%
    mutate(Batch = row$Batch)
}

# Execution (commented out as it depends on external files)
# appended_data <- df_names %>%
#   split(1:nrow(.)) %>%
#   map_df(read_and_sample)

# For demo purposes, we assume 'appended_data' is created here
```

## Summarise the data by binning

We calculate the cylinder volume and log-transform it to create bins.

```{r}
# Example logic for calculation and binning
process_data <- function(df) {
  df %>%
    mutate(
      cyl_vol = pi * `Length (µm)` * (`Width (µm)` / 2)^2,
      log_cyl_vol = log10(cyl_vol)
    ) %>%
    # Create 40 bins between 0 and 10
    mutate(bins = cut(log_cyl_vol, breaks = seq(0, 10, length.out = 41), include.lowest = TRUE)) %>%
    group_by(Batch, bins) %>%
    summarise(
      Count = n(),
      Sum_cyl_vol = sum(cyl_vol, na.rm = TRUE),
      across(c(Circularity, Convexity, Solidity, `Aspect Ratio`), median, na.rm = TRUE),
      .groups = "drop"
    )
}
```

## Create Excel output for Spotfire

We calculate percentages per batch for both volume and count.

```{r}
finalize_for_export <- function(df_summ) {
  df_summ %>%
    group_by(Batch) %>%
    mutate(
      perc_cyl_vol = 100 * Sum_cyl_vol / sum(Sum_cyl_vol),
      cum_perc_cyl_vol = cumsum(perc_cyl_vol),
      perc_N = 100 * Count / sum(Count),
      cum_perc_N = cumsum(perc_N)
    ) %>%
    ungroup()
}

# write_xlsx(df_final, "df_binned_data.xlsx")
```

## Create pivoted data for modelling

Finally, we reshape the data to a wide format, which is often required by modelling software like MODDE.

```{r}
create_pivoted_data <- function(df_final) {
  df_final %>%
    pivot_wider(
      id_cols = Batch,
      names_from = bins,
      values_from = c(Count, Circularity, Convexity),
      names_glue = "{.value}_{bins}"
    )
}

# write_xlsx(df_pivoted, "df_pivoted_data.xlsx")
```

# Summary

This workflow demonstrates how to:
1.  Batch-load and sample large text files using `purrr` and `readr`.
2.  Perform custom geometric calculations with `dplyr`.
3.  Bin continuous data and calculate grouped statistics.
4.  Reshape data for external modelling tools using `tidyr`.
