---
title: "Text and Annotation in R"
output: html_document
---

# Text and Annotation

Adding text and annotations can provide crucial context to a plot. In `ggplot2`, we use `annotate()`, `geom_text()`, and `geom_label()`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Global Annotations

`annotate()` allows you to add text, rectangles, or lines that aren't mapped to the data frame.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  annotate("text", x = 5, y = 40, label = "High Efficiency ->", color = "red", fontface = "bold") +
  annotate("rect", xmin = 5, xmax = 7, ymin = 10, ymax = 20, alpha = 0.2, fill = "blue") +
  theme_minimal()
```

## Data-Driven Text with `geom_text()`

If you want to label specific points from your data, use `geom_text()`.

```{r}
# Label only the most efficient cars
labels_df <- mpg %>% filter(hwy > 40)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  geom_text(data = labels_df, aes(label = model), vjust = -1, color = "darkblue") +
  theme_bw()
```

## Labels with Backgrounds: `geom_label()`

`geom_label()` is similar to `geom_text()` but draws a box behind the text for better readability.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.3) +
  geom_label(data = labels_df, aes(label = model), fill = "white") +
  theme_classic()
```

## Summary

-   **Contextual Text:** Use `annotate("text", ...)` for arbitrary labels.
-   **Data Labels:** Use `geom_text()` or `geom_label()` to map text to data points.
-   **Highlighting:** Use `annotate("rect", ...)` or `geom_vline()` / `geom_hline()` to highlight specific regions or values.
