---
title: "Introduction to R Visualization with ggplot2"
output: html_document
---

# Introduction to Data Visualization in R

Visualization is one of the core strengths of the R language. While base R provides many plotting functions, the `tidyverse` ecosystem—and specifically the `ggplot2` package—has become the industry standard for creating publication-quality graphics.

In this section, we adapt concepts from the "Python Data Science Handbook" to R. The "Grammar of Graphics" used by `ggplot2` allows us to describe and build complex visualizations by layering components.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

### Core Concepts

1.  **Aesthetics (aes):** Mapping data variables to visual properties (x, y, color, size).
2.  **Geometries (geoms):** Defining the type of plot (points, lines, bars).
3.  **Facets:** Splitting a plot into multiple subplots based on a factor.
4.  **Themes:** Controlling the non-data visual elements (fonts, background).

Throughout these supplementary notebooks, we will explore:
-   Simple Line and Scatter Plots
-   Histograms and Density Plots
-   Customizing Legends and Colorbars
-   Advanced Layouts and Annotations

Let's begin with a simple example:

```{r}
# Mapping 'mpg' data: Engine displacement vs Highway miles per gallon
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(title = "Fuel Efficiency vs Engine Size", x = "Engine Displacement (L)", y = "Highway MPG") +
  theme_minimal()
```
