---
title: "Multiple Subplots and Layouts in R"
output: html_document
---

# Multiple Subplots and Layouts

In R, there are several ways to arrange multiple plots. The most idiomatic way is through **Faceting** (splitting one plot by a variable) or using layout packages like `ggpubr` or `patchwork`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)
```

## Faceting: Subplots by Data

Faceting automatically creates a grid of subplots based on a categorical variable.

```{r}
# Using the built-in mpg dataset
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~class, ncol = 3) +
  theme_bw()
```

## Arranging Independent Plots

If you have separate plot objects, you can arrange them using `ggarrange` from the `ggpubr` package.

```{r}
p1 <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() + theme_minimal()
p2 <- ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) + geom_boxplot() + theme_minimal()
p3 <- ggplot(mpg, aes(x = hwy)) + geom_histogram(bins = 20) + theme_minimal()

# Arrange in a grid
ggarrange(p1, p2, p3, ncol = 2, nrow = 2, labels = c("A", "B", "C"))
```

## Grid Layouts with `facet_grid`

`facet_grid` is useful when you want to split subplots by two variables.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl) +
  theme_light()
```

## Summary

-   **Faceting:** Use `facet_wrap()` for one variable and `facet_grid()` for two.
-   **Layouts:** Use `ggpubr::ggarrange()` to combine different types of plots into one figure.
-   **Consistency:** Subplots created via faceting automatically share axes and scales, ensuring visual consistency.
