---
title: "Customizing Colorbars in R"
output: html_document
---

# Customizing Colorbars

In R, continuous color scales automatically produce a colorbar (legend). We can customize these using the `scale_*_gradient*` family of functions and `guides()`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Basic Colorbars

Colorbars are generated when a continuous variable is mapped to the `color` or `fill` aesthetic.

```{r}
# Using the built-in faithful dataset
ggplot(faithful, aes(x = eruptions, y = waiting, color = eruptions)) +
  geom_point() +
  scale_color_viridis_c() +
  theme_minimal()
```

## Customizing Color Scales

We can control the limits, colors, and title of the colorbar.

```{r}
ggplot(faithful, aes(x = eruptions, y = waiting, color = eruptions)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red") +
  labs(color = "Eruption Duration (min)") +
  theme_bw()
```

## Controlling Colorbar Appearance with `guides()`

Use `guide_colorbar()` for fine-grained control over the colorbar's dimensions and labels.

```{r}
ggplot(faithful, aes(x = eruptions, y = waiting, color = eruptions)) +
  geom_point() +
  scale_color_viridis_c() +
  guides(color = guide_colorbar(barwidth = 15, barheight = 1, title.position = "top")) +
  theme(legend.position = "bottom")
```

## Summary

-   **Continuous Mapping:** Continuous variables mapped to color/fill produce colorbars.
-   **Color Palettes:** Use `scale_color_viridis_c()` or `scale_color_distiller()` for pre-defined palettes.
-   **Manual Control:** Use `scale_color_gradient()` for custom start/end colors.
-   **Refinement:** Use `guide_colorbar()` within `guides()` to adjust physical properties.
