---
title: "Visualizing Errors in R"
output: html_document
---

# Visualizing Errors

For any scientific measurement, accurate visualization of error is just as important as the visualization of the data itself. In `ggplot2`, we have several geoms to represent uncertainty.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Basic Errorbars

We can add error bars to discrete points using `geom_errorbar()`.

```{r}
# Generate some data with errors
x <- seq(0, 10, length.out = 30)
dy <- 0.8
y <- sin(x) + dy * rnorm(30)

df <- tibble(x = x, y = y, dy = dy)

ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_errorbar(aes(ymin = y - dy, ymax = y + dy), width = 0.2) +
  theme_minimal()
```

## Continuous Errors: Ribbons

When visualizing errors for continuous functions, a "ribbon" (equivalent to Matplotlib's `fill_between`) is often more effective than individual error bars.

```{r}
# Generate a smooth fit with uncertainty
df_fit <- tibble(
  x = seq(0, 10, length.out = 100),
  y = sin(x),
  error = 0.2
)

ggplot(df_fit, aes(x = x, y = y)) +
  geom_line(color = "red") +
  geom_ribbon(aes(ymin = y - error, ymax = y + error), fill = "red", alpha = 0.2) +
  labs(title = "Line Plot with Error Ribbon") +
  theme_bw()
```

## Summary

-   **Discrete Points:** Use `geom_errorbar()` or `geom_pointrange()`.
-   **Continuous Functions:** Use `geom_ribbon()` to show a shaded region of uncertainty.
-   **Aesthetics:** Always map `ymin` and `ymax` to your upper and lower bounds.
