---
title: "Histograms, Binnings, and Density in R"
output: html_document
---

# Histograms, Binnings, and Density

A simple histogram can be a great first step in understanding a dataset. In R, we use `ggplot2` for both 1D and 2D distribution analysis.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Basic Histograms

In `ggplot2`, we use `geom_histogram()` to visualize the distribution of a continuous variable.

```{r}
set.seed(0)
df <- tibble(data = rnorm(1000))

ggplot(df, aes(x = data)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal()
```

## Comparing Multiple Distributions

We can compare multiple distributions by mapping a grouping variable to the `fill` aesthetic and using transparency (`alpha`).

```{r}
df_multi <- tibble(
  group = rep(c("A", "B", "C"), each = 1000),
  value = c(rnorm(1000, 0, 0.8), rnorm(1000, -2, 1), rnorm(1000, 3, 2))
)

ggplot(df_multi, aes(x = value, fill = group)) +
  geom_histogram(bins = 40, alpha = 0.4, position = "identity") +
  theme_classic()
```

## Density Plots (KDE)

`geom_density()` provides a smooth estimate of the distribution, equivalent to Kernel Density Estimation.

```{r}
ggplot(df_multi, aes(x = value, fill = group)) +
  geom_density(alpha = 0.3) +
  theme_minimal()
```

## Two-Dimensional Histograms and Binnings

Just as we create histograms in one dimension, we can also create them in two dimensions.

```{r}
# Generate multivariate normal data
library(MASS)
set.seed(42)
mu <- c(0, 0)
Sigma <- matrix(c(1, 1, 1, 2), 2)
df_2d <- as_tibble(mvrnorm(10000, mu, Sigma))
colnames(df_2d) <- c("x", "y")

# 2D Histogram
ggplot(df_2d, aes(x = x, y = y)) +
  geom_bin2d(bins = 30) +
  scale_fill_viridis_c() +
  theme_bw()
```

### Hexagonal Binnings

Hexagonal bins are often preferred for 2D distributions as they minimize visual artifacts.

```{r}
ggplot(df_2d, aes(x = x, y = y)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c() +
  theme_minimal()
```

### 2D Density Contours

```{r}
ggplot(df_2d, aes(x = x, y = y)) +
  geom_density_2d_filled() +
  theme_minimal()
```

## Summary

-   **1D Distributions:** Use `geom_histogram()` or `geom_density()`.
-   **2D Distributions:** Use `geom_bin2d()`, `geom_hex()`, or `geom_density_2d()`.
-   **Tuning:** Use `bins`, `binwidth`, and `alpha` to control the granularity and visibility of your plots.
