---
title: "Density and Contour Plots in R"
output: html_document
---

# Density and Contour Plots

Sometimes it is useful to display three-dimensional data in two dimensions using contours or color-coded regions. In `ggplot2`, we can use `geom_contour()`, `geom_density_2d()`, and `geom_raster()`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Visualizing a 3D Function

We can visualize a function $z = f(x, y)$ by creating a grid of $x$ and $y$ values.

```{r}
# Define a function
f <- function(x, y) {
  sin(x)^10 + cos(10 + y * x) * cos(x)
}

# Create a grid
grid <- expand_grid(
  x = seq(0, 5, length.out = 100),
  y = seq(0, 5, length.out = 100)
) %>%
  mutate(z = f(x, y))

# Contour plot
ggplot(grid, aes(x = x, y = y, z = z)) +
  geom_contour(aes(color = after_stat(level))) +
  scale_color_viridis_c() +
  theme_minimal()
```

## Filled Contours and Rasters

For a more continuous look, we can use `geom_raster()` (which is highly efficient for grids) or `geom_contour_filled()`.

```{r}
# Raster plot (Heatmap style)
ggplot(grid, aes(x = x, y = y, fill = z)) +
  geom_raster() +
  scale_fill_viridis_c() +
  theme_void()

# Filled contours
ggplot(grid, aes(x = x, y = y, z = z)) +
  geom_contour_filled() +
  labs(title = "Filled Contour Plot") +
  theme_minimal()
```

## 2D Density Estimation

If we have a collection of points, we can estimate their 2D density.

```{r}
# Generate points
set.seed(42)
df_points <- tibble(
  x = rnorm(1000),
  y = rnorm(1000)
)

ggplot(df_points, aes(x = x, y = y)) +
  geom_point(alpha = 0.1) +
  geom_density_2d(color = "red") +
  theme_bw()
```

## Summary

-   **Gridded Data:** Use `geom_contour()` or `geom_raster()`.
-   **Point Data:** Use `geom_density_2d()` or `geom_density_2d_filled()`.
-   **Color Scales:** `scale_fill_viridis_c()` is often a good choice for visibility.
