---
title: "Simple Scatter Plots in R"
output: html_document
---

# Simple Scatter Plots

Another commonly used plot type is the simple scatter plot, a close cousin of the line plot. Instead of points being joined by line segments, here the points are represented individually with a dot, circle, or other shape.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Scatter Plots with `geom_point()`

In `ggplot2`, scatter plots are created using the `geom_point()` geometry.

```{r}
# Generate data
df <- tibble(
  x = seq(0, 10, length.out = 30),
  y = sin(x)
)

# Basic scatter plot
ggplot(df, aes(x = x, y = y)) +
  geom_point()
```

## Customizing Markers

We can change the shape and size of the points globally within `geom_point()`.

```{r}
ggplot(df, aes(x = x, y = y)) +
  geom_point(shape = 17, size = 3, color = "darkred") +
  theme_minimal()
```

R provides 25 built-in shapes, referenced by numeric codes (e.g., 0-25).

## Mapping Aesthetics to Data

The real power of scatter plots comes when we map visual properties to variables in our data.

```{r}
set.seed(0)
random_data <- tibble(
  x = rnorm(100),
  y = rnorm(100),
  colors = runif(100),
  sizes = runif(100) * 10
)

ggplot(random_data, aes(x = x, y = y, color = colors, size = sizes)) +
  geom_point(alpha = 0.5) +
  scale_color_viridis_c() +
  theme_minimal()
```

### Multidimensional Exploration with Iris

Using the built-in `iris` dataset, we can visualize four dimensions simultaneously: sepal length, sepal width, petal width (size), and species (color).

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, size = Petal.Width)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Iris Data: Multiple Dimensions",
    x = "Sepal Length",
    y = "Sepal Width"
  ) +
  theme_bw()
```

## Summary

-   **Geometries:** Use `geom_point()` for scatter plots.
-   **Aesthetics:** Use `aes()` to map variables to `color`, `size`, and `shape`.
-   **Customization:** Fixed values for color/size should go outside `aes()`, while variable mappings go inside.
