---
title: "Simple Line Plots in R"
output: html_document
---

# Simple Line Plots

Perhaps the simplest of all plots is the visualization of a single function $y = f(x)$. Here we will create simple line plots using `ggplot2`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Creating a Simple Plot

In `ggplot2`, we typically start with a data frame and map our variables to aesthetics. For functions, we can generate a data frame of $x$ and $y$ values.

```{r}
# Generate data
df <- tibble(
  x = seq(0, 10, length.out = 1000),
  y = sin(x)
)

# Basic line plot
ggplot(df, aes(x = x, y = y)) +
  geom_line()
```

## Adjusting the Plot: Line Colors and Styles

We can control the appearance of the line using arguments within `geom_line()`.

```{r}
# Color and Linetype
ggplot(df, aes(x = x, y = y)) +
  geom_line(color = "blue", linetype = "dashed", size = 1) +
  theme_minimal()
```

Common linetypes in R include: `"solid"`, `"dashed"`, `"dotted"`, `"dotdash"`, `"longdash"`, and `"twodash"`.

## Multiple Lines

To plot multiple functions, we can either layer geoms or, more idiomatically, "tidy" our data.

```{r}
# Layering approach
ggplot(df, aes(x = x)) +
  geom_line(aes(y = sin(x), color = "Sine")) +
  geom_line(aes(y = cos(x), color = "Cosine")) +
  labs(color = "Function")
```

## Adjusting Axes and Labeling

`ggplot2` provides several functions for controlling axis limits and labels.

```{r}
ggplot(df, aes(x = x, y = y)) +
  geom_line() +
  xlim(-1, 11) +
  ylim(-1.5, 1.5) +
  labs(
    title = "A Sine Curve",
    x = "x",
    y = "sin(x)"
  ) +
  theme_bw()
```

## Summary

-   **Data First:** `ggplot2` works best when data is in a data frame (tibble).
-   **Layers:** Use `+` to add `geom_line()`, `labs()`, and `theme()` layers.
-   **Scales:** Use `xlim()` and `ylim()` for simple coordinate adjustments.
