---
title: "Three-Dimensional Plotting in R"
output: html_document
---

# Three-Dimensional Plotting

While 2D plots are standard, sometimes 3D visualization is necessary to explore complex relationships. In R, we often use `plotly` for interactive 3D plots or base R's `persp()`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
```

## 3.1 3D Scatter Plots

With `plotly`, we can create interactive 3D scatter plots where you can rotate and zoom to explore the data.

```{r}
# Using the built-in iris dataset
plot_ly(iris, x = ~Sepal.Length, y = ~Sepal.Width, z = ~Petal.Length,
        color = ~Species, type = "scatter3d", mode = "markers")
```

## 3.2 3D Surface Plots

We can visualize a mathematical function or a grid of data as a 3D surface.

```{r}
# Define a function
f <- function(x, y) {
  sin(sqrt(x^2 + y^2))
}

# Create a grid
x <- seq(-8, 8, length.out = 50)
y <- seq(-8, 8, length.out = 50)
z <- outer(x, y, f)

# Surface plot
plot_ly(z = ~z, type = "surface")
```

## Summary

-   **Interactive Exploration:** `plotly` is the go-to package for interactive 3D visualization in R.
-   **Static 3D:** Packages like `plot3D` or base R's `persp()` can be used for static publications.
-   **Caution:** 3D plots can be difficult to interpret; always consider if a 2D faceted plot or heatmap would convey the information more clearly.
