---
title: "Customizing Ticks and Scales in R"
output: html_document
---

# Customizing Ticks and Scales

Matplotlib's tick locators and formatters have direct equivalents in R through the `scale_*` functions in `ggplot2` and the `scales` package.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales) # Essential for tick formatting
```

## Basic Tick Customization

We can control where ticks appear using the `breaks` argument and how they are labeled using `labels`.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  scale_x_continuous(breaks = seq(1, 7, by = 1)) +
  scale_y_continuous(breaks = c(10, 25, 40), labels = c("Low", "Medium", "High")) +
  theme_minimal()
```

## Logarithmic Scales

Log scales are common in bioinformatics and data science. In `ggplot2`, we use `scale_*_log10()`.

```{r}
ggplot(msleep, aes(x = bodywt, y = brainwt)) +
  geom_point() +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = label_log()) +
  theme_bw()
```

## Hiding Ticks and Labels

To remove ticks or labels, we use `theme()`.

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar() +
  theme(
    axis.text.x = element_blank(), # Hide x-axis text
    axis.ticks.x = element_blank() # Hide x-axis ticks
  )
```

## Summary

-   **Manual Control:** Use `breaks` and `labels` inside `scale_x_continuous()` or similar.
-   **Special Formats:** Use the `scales` package (e.g., `label_percent()`, `label_dollar()`, `label_log()`) for consistent formatting.
-   **Cosmetic Control:** Use `theme()` to adjust the appearance or visibility of ticks and text.
