---
title: "Geographic Data in R"
output: html_document
---

# Geographic Data and Mapping

Geographic visualization is a common task in data science. In R, the `sf` (Simple Features) package is the modern standard for handling spatial data, while `ggplot2` and `leaflet` are used for visualization.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(maps)
library(leaflet)
```

## Static Maps with `ggplot2`

`ggplot2` provides `geom_sf()` for plotting spatial objects.

```{r}
# Load world map data
world <- st_as_sf(map('world', plot = FALSE, fill = TRUE))

# Plot world map
ggplot(data = world) +
  geom_sf(fill = "antiquewhite") +
  coord_sf(expand = FALSE) +
  theme_minimal()
```

## Plotting Points on a Map

We can overlay data points (like cities) onto a map.

```{r}
# Example cities
cities <- tibble(
  city = c("London", "New York", "Tokyo"),
  lat = c(51.5074, 40.7128, 35.6895),
  lon = c(-0.1278, -74.0060, 139.6917)
)

ggplot(data = world) +
  geom_sf() +
  geom_point(data = cities, aes(x = lon, y = lat), color = "red", size = 3) +
  theme_bw()
```

## Interactive Maps with `leaflet`

For web-based projects, `leaflet` provides highly interactive maps.

```{r}
leaflet(cities) %>%
  addTiles() %>%
  addMarkers(~lon, ~lat, popup = ~city)
```

## Summary

-   **Data Standard:** Use the `sf` package to handle spatial data.
-   **Static Maps:** Use `geom_sf()` in `ggplot2` for reproducible static figures.
-   **Interactive Maps:** Use `leaflet` for interactive, zoomable map widgets.
