---
title: "Advanced Data Visualization with ggplot2"
output: html_document
---

# Advanced Visualization in R

While base R has useful plotting tools, `ggplot2` (part of the `tidyverse`) is the standard for sophisticated statistical visualization. It is the R equivalent to Python's `seaborn` and `plotnine`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)
library(GGally) # For pair plots
```

## Exploring Statistical Plots

`ggplot2` provides high-level commands to create a variety of plot types useful for statistical data exploration.

### Histograms and Densities

```{r}
# Using the built-in iris dataset
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
  theme_minimal()

# Density plot
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.3) +
  theme_minimal()
```

### Pair Plots

In R, the `GGally` package provides a powerful equivalent to Seaborn's `pairplot`.

```{r, message=FALSE}
ggpairs(iris, aes(color = Species, alpha = 0.5))
```

### Faceted Plots

`ggplot2` makes faceting extremely simple. We'll look at the `tips` dataset (available in many R packages or via simulation).

```{r}
# Loading a similar dataset or using a simulated one for demo
data(tips, package = "reshape2")

ggplot(tips, aes(x = total_bill, y = tip)) +
  geom_point(alpha = 0.5) +
  facet_grid(sex ~ time) +
  theme_bw()
```

### Violin Plots

A great way to compare distributions across categories.

```{r}
ggplot(tips, aes(x = day, y = total_bill, fill = sex)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  theme_classic()
```

### Joint Distributions and Regression

`geom_smooth()` automatically fits regression models to the data.

```{r}
ggplot(tips, aes(x = total_bill, y = tip)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "Joint Distribution with Regression Line") +
  theme_minimal()
```

## Summary

-   `ggplot2` provides a consistent "Grammar of Graphics" for building complex plots.
-   `GGally` extends this for multivariate exploration (pair plots).
-   Faceting and layering allow for deep data exploration with minimal code.
