---
title: "Themes and Settings in R"
output: html_document
---

# Themes and Settings

In R, global plot appearance is controlled by **Themes**. `ggplot2` comes with several built-in themes and allows you to create or modify your own.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Built-in Themes

`ggplot2` provides several complete themes.

```{r}
p <- ggplot(mpg, aes(x = displ, y = hwy, color = drv)) + geom_point()

# Try different themes
p + theme_grey()    # The default
p + theme_bw()      # Black and white
p + theme_minimal() # Clean and modern
p + theme_classic() # No gridlines
```

## Setting a Global Theme

You can set a theme for all plots in a session using `theme_set()`.

```{r}
theme_set(theme_light())

# All subsequent plots will use theme_light
ggplot(mpg, aes(x = hwy)) + geom_histogram(bins = 30)
```

## Customizing Theme Elements

The `theme()` function allows you to modify specific visual elements of a plot.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  theme(
    panel.grid.major = element_line(color = "red", linetype = "dotted"),
    axis.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )
```

## Summary

-   **Complete Themes:** Start with `theme_minimal()`, `theme_bw()`, etc.
-   **Global Settings:** Use `theme_set()` at the start of your script.
-   **Fine-Grained Control:** Use `theme()` with `element_text()`, `element_line()`, and `element_rect()` to customize every aspect of the plot.
