---
title: "Week 1 - Data handling with the Tidyverse"
output: html_document
---

# Week 1 - Data handling

In R, the `tidyverse` suite of packages (including `dplyr`, `tidyr`, and `readr`) provides powerful tools to handle datasets and apply basic operations on them. These packages are the R equivalents to Python's `pandas` and `numpy`.

The `tidyverse` offers a consistent system for manipulating data - see the [official documentation](https://www.tidyverse.org/).

### Aims

- Gain familiarity with data handling using the `tidyverse`:
    - Create, read, and write data.
    - Select a subset of variables (columns).
    - Filter rows based on their values.
    - Sort datasets.
    - Create new columns or modify existing ones.
    - Summarise data.
    - Handle missing data.
    - Merge datasets.

### Loading the Tidyverse

We extend R's base functionality by loading the `tidyverse` library.

```{r setup, message=FALSE}
library(tidyverse)
```

### Reading datasets

We are going to use the METABRIC dataset `metabric_clinical_and_expression_data.csv` containing information about breast cancer patients.

To read a CSV file, we use the `read_csv()` function from the `readr` package:

```{r}
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv")
```

```{r}
metabric
```

Generally, rows in a `tibble` (the tidyverse version of a data frame) are the **observations** (patients) whereas columns are the observed **variables** (Cohort, Age_at_diagnosis, etc.).

### Exploring data

The `tibble` is a 2-dimensional table where columns can contain different data types (e.g., character, integer, double, logical).

Using the `glimpse()` function, we can view basic information about our data:

```{r}
glimpse(metabric)
```

We can access dimensions using `dim()` and column names using `colnames()`:

```{r}
dim(metabric)
colnames(metabric)
```

The `summary()` function computes summary statistics for all columns:

```{r}
summary(metabric)
```

To access a specific column as a vector, use the `$` operator:

```{r}
mean(metabric$Survival_time, na.rm = TRUE)
```

### Selecting columns and rows

In the `tidyverse`, we use `select()` to pick columns and `slice()` or `filter()` to pick rows.

```{r}
# Select the first three rows
metabric %>% slice(1:3)

# Select specific columns
metabric %>% select(Cohort, Chemotherapy) %>% head(3)
```

### Filtering rows

You can choose rows that match criteria using `filter()`.

```{r}
# Filter for patients who died of the disease
metabric %>% filter(Vital_status == "Died of Disease") %>% head()

# Multiple conditions
metabric %>%
  filter(Vital_status == "Died of Disease", Age_at_diagnosis > 70) %>%
  head()
```

### Defining new columns

Use `mutate()` to create or transform columns.

```{r}
metabric <- metabric %>%
  mutate(Age_today = Age_at_diagnosis + Survival_time/365)

metabric %>% select(Age_at_diagnosis, Survival_time, Age_today) %>% head()
```

### Sort data

Use `arrange()` to sort. Wrap a column in `desc()` for descending order.

```{r}
metabric %>% arrange(desc(Tumour_size)) %>% head()
```

### Missing data

R uses `NA` to represent missing data.

```{r}
# Count NAs per column
colSums(is.na(metabric))

# Drop rows with any NAs
metabric_clean <- metabric %>% drop_na()
```

### Grouping and Summarising

```{r}
metabric %>%
  group_by(Cohort) %>%
  summarise(mean_survival = mean(Survival_time, na.rm = TRUE))
```

### Merge datasets

Use `bind_rows()` to concatenate and `left_join()` or `inner_join()` to merge based on keys.

```{r}
df1 <- tibble(Patient_ID = c("A", "B"), Val1 = c(1, 2))
df2 <- tibble(Patient_ID = c("A", "B"), Val2 = c(3, 4))

left_join(df1, df2, by = "Patient_ID")
```

## Assignment

1.  Read `metabric_clinical_and_expression_data.csv` and export its summary statistics to a CSV and Excel file.
2.  Calculate the mean tumour size of patients grouped by vital status and tumour stage.
3.  Identify patient IDs with missing tumour size data.
4.  (Bonus) Try these operations on your own dataset.
