---
title: "Week 4 - Statistical Tests and Modelling"
output: html_document
---

# Statistical Analysis in R

In this session, we will explore how to carry out statistical tests and modelling in R. R is a language built for statistics, so many of these tools are included in the base installation (`stats` package), and we will use the `tidyverse` for data handling and `broom` for tidying model outputs.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(ggpubr) # for easy statistical annotations on plots
```

# Data
We will use the METABRIC dataset. Many statistical functions in R handle missing data using the `na.rm = TRUE` argument, but for modelling, we will use a version of the dataset with NAs removed.

```{r}
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv") %>%
  drop_na()

summary(metabric)
```

# Statistical tests

## Tests for normality
Before choosing a test (e.g., t-test vs Mann-Whitney), we often check for normality. We can use the **Shapiro-Wilk test** or the **Kolmogorov-Smirnov test**.

```{r}
ggplot(metabric, aes(x = Age_at_diagnosis)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal()

# Shapiro-Wilk test (note: usually limited to sample sizes < 5000)
shapiro.test(metabric$Age_at_diagnosis[1:1000])
```

The low p-value suggests the data is non-normal.

## Correlation
To test the relationship between two continuous variables:

```{r}
# Pearson correlation (for normal data)
cor.test(metabric$ESR1, metabric$GATA3, method = "pearson")

# Spearman rank correlation (for non-normal data)
cor.test(metabric$Tumour_size, metabric$Mutation_count, method = "spearman")
```

## T-test and Mann-Whitney U test
To compare means between two groups (e.g., ER status):

```{r}
# T-test (parametric)
t.test(Age_at_diagnosis ~ ER_status, data = metabric)

# Wilcoxon / Mann-Whitney U test (non-parametric)
wilcox.test(Survival_time ~ ER_status, data = metabric)
```

## ANOVA and Kruskal-Wallis
To compare >2 groups:

```{r}
# ANOVA
res_aov <- aov(Survival_time ~ Cancer_type, data = metabric)
summary(res_aov)

# Kruskal-Wallis (non-parametric)
kruskal.test(Survival_time ~ Cancer_type, data = metabric)
```

## Chi-square test
For categorical variables:

```{r}
# Create contingency table
cont_table <- table(metabric$ER_status, metabric$PR_status)
cont_table

# Chi-square test
chisq.test(cont_table)
```

# Data Transformation
Transforming skewed data (like tumour size) can make it more suitable for modelling.

```{r}
metabric <- metabric %>%
  mutate(
    Tumour_size_log10 = log10(Tumour_size),
    Tumour_size_sqrt  = sqrt(Tumour_size)
  )

# Visualizing transformations
metabric %>%
  select(Tumour_size, Tumour_size_log10, Tumour_size_sqrt) %>%
  pivot_longer(everything(), names_to = "Transformation", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~Transformation, scales = "free") +
  theme_bw()
```

# Modelling

## Simple Linear Regression
Predicting survival time based on tumour size.

```{r}
# Fit model
simple_model <- lm(Survival_time ~ Tumour_size, data = metabric)

# Display results
summary(simple_model)

# Use broom to get a tidy table
tidy(simple_model)
glance(simple_model)
```

## Multivariate Linear Regression
Considering multiple predictors simultaneously.

```{r}
complex_model <- lm(Survival_time ~ Tumour_size + Tumour_stage + Cancer_type + ER_status, data = metabric)
summary(complex_model)
```

# Exercises

## Exercise 1
Is there a significant difference between the tumour size of patients who received chemotherapy versus those who did not? Visualize and test.

## Exercise 2
Test the correlation between tumour size and survival time. Compare the correlation coefficient for raw vs log10-transformed data.

## Exercise 3
Investigate if Chemotherapy and Radiotherapy treatments are independent using a Chi-square test.
