---
title: "Week 0 - Basic R concepts and functions"
output: html_document
---

# Week 0 - Basic R concepts and functions

### Aims

- To recapitulate R's basic functionality (syntax, data types, structures, conditioning and I/O)
- To learn and/or practice writing functions in R

## Basic R

### Printing

In R, we use `print()` or `cat()` to output text to the console.

```{r}
print('Hello R!') # to print some text, enclose it between single quotation marks
print("I'm here today!")    # or double quotation marks
print(34)                   # print an integer
print(2 + 4)                # print the result of an arithmetic operation
# cat() is often better for combining text and numbers
cat("The answer is", 42, "\n")
```

You can include a comment in R by prefixing some text with a `#` character. All text following the `#` will then be ignored by the interpreter.

### Variables

A variable can be assigned to a simple value or the outcome of a more complex expression. The `<-` operator is the standard assignment operator in R.

```{r}
x <- 3     # assignment of a simple value
print(x)
y <- x + 5 # assignment of a more complex expression
print(y)
i <- 12
print(i)
i <- i + 1 # incrementing
print(i)
```

### Simple data types

R has several main data types: numeric (includes integer and double), character (strings), logical (boolean), and complex.

```{r}
a <- 2           # numeric
b <- 5.0         # numeric
c <- 'word'      # character
d <- 4 > 5       # logical (TRUE or FALSE)
e <- NULL        # special value for "nothing"
print(class(a))  # check the class of a variable
```

### Arithmetic operations

```{r}
3 + 2             # addition
3 - 2             # subtraction
3 * 2             # multiplication
3 / 2             # division
3 %/% 2           # integer division
3 %% 2            # remainder (modulo)
3 ^ 2             # exponent
```

### Data structures

R has several core data structures to store the data types shown above:

- Vectors (Atomic)
- Lists
- Matrices and Arrays
- Data Frames and Tibbles
- Factors

##### Vectors

A vector is an ordered collection of elements of the same type.

```{r}
a <- c('red', 'blue', 'green', "yellow") # c() combines elements
length(a)                                # length
d <- a[1]                                # access first element (R is 1-indexed!)
e <- a[2:3]                              # access a slice
f <- a[length(a)]                        # access last element
a[2] <- "purple"                         # modify an element
g <- c('re', 'bl', 'gr')                 # concatenation is done with c()
'red' %in% a                             # check existence
sort(a)                                  # sort
```

##### Lists

Lists can contain elements of different types and lengths.

```{r}
a <- list(number = 1, text = "apple", flag = TRUE)
a[[2]] # access by index
a$text # access by name
```

### Working with strings

In R, strings are character vectors. The `stringr` package (part of tidyverse) is highly recommended.

```{r}
library(stringr)
a <- 'red'
char <- substr(a, 1, 1)            # access substring
b <- paste0('red', 'blue')         # concatenation
c <- str_split('1, 2, three', ',') # split
d <- str_c(c('1', '2', 'three'), collapse = ".") # join
```

### Conditions

```{r}
a <- 1; b <- 0
if (a + b == 3) {
    print('Three')
} else if (a + b == 1) {
    print('One')
} else {
    print('?')
}
```

### Loops

##### for

```{r}
colors <- c('red', 'blue', 'green')
for (color in colors) {
    print(color)
}
```

##### while

```{r}
number <- 1
while (number < 10) {
    print(number)
    number <- number + 1
}
```

### Reading and writing files

We typically use `read_csv()` and `write_csv()` the `readr` package.

```{r, eval=FALSE}
library(readr)
# reading file
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv")

# writing to a file
write_lines(c("I love programming in R!", "I love making scripts."), "programming.txt")
```

### Getting help

```{r, eval=FALSE}
help(length)
?length
```

## Functions

### Overview

A function is a block of code that only runs when called. In R, we define functions using the `function` keyword.

```{r}
print_text <- function(input_text, verbose = FALSE) {
    if (verbose) {
        cat("Processing input...\n")
    }
    return(input_text)
}

result <- print_text("Hello world", verbose = TRUE)
print(result)
```

## External functions and Packages

### Loading Packages

In R, we extend functionality by loading **packages** using the `library()` command.

```{r}
# Example
# library(tidyverse)
```

Once a package is loaded, its functions are available. You can also call a function without loading the package using the `::` operator (e.g., `dplyr::filter()`).

## Assignment

1. Use the METABRIC dataset to answer the following questions:
    - How many unique patients?
    - How many older than 75 at diagnosis?
    - What is the age range at diagnosis?

2. Write a function that finds the most abundant amino acid in a protein sequence.

3. Debug the provided function (Note: In R, arguments with defaults should follow those without).

4. Find the most frequent word in "imagine_lyrics.txt" using R functions.
