---
title: "Week 6 - An introduction to Machine Learning (PART 2)"
output: html_document
---

# Week 6 - An introduction to Machine Learning (PART 2)

## Motivation

> In this session, we transition regression to **classification**. We will explore Decision Trees and Random Forests, focusing on model selection, cross-validation, and hyperparameter tuning using the `tidymodels` framework.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(ranger)
library(vip) # For feature importance
library(palmerpenguins) # For a classification example
```

## 1. Classification with Decision Trees

We will use the `penguins` dataset to predict species.

```{r}
# Data Preparation
data("penguins")
penguins_clean <- penguins %>% drop_na()

# Split
set.seed(123)
penguin_split <- initial_split(penguins_clean, prop = 0.7)
penguin_train <- training(penguin_split)
penguin_test  <- testing(penguin_split)

# Model Spec
tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

# Fit
tree_fit <- tree_spec %>%
  fit(species ~ bill_length_mm + bill_depth_mm + flipper_length_mm, data = penguin_train)

# Predict
penguin_results <- penguin_test %>%
  select(species) %>%
  bind_cols(predict(tree_fit, penguin_test))

# Metrics
conf_mat(penguin_results, truth = species, estimate = .pred_class)
accuracy(penguin_results, truth = species, estimate = .pred_class)
```

## 2. Random Forests and Feature Importance

Random Forests improve performance by averaging multiple trees.

```{r}
rf_spec <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_fit <- rf_spec %>%
  fit(species ~ ., data = penguin_train)

# Variable Importance
vip(rf_fit)
```

## 3. Cross-Validation and Hyperparameter Tuning

How do we pick the best parameters? We use `tune()` and grid search.

```{r}
# Tuneable spec
tune_spec <- rand_forest(
  mtry = tune(),
  trees = 100,
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Recipe
penguin_recipe <- recipe(species ~ ., data = penguin_train) %>%
  step_dummy(all_nominal_predictors())

# Workflow
penguin_wf <- workflow() %>%
  add_recipe(penguin_recipe) %>%
  add_model(tune_spec)

# Tuning
set.seed(456)
penguin_folds <- vfold_cv(penguin_train, v = 5)

tune_res <- tune_grid(
  penguin_wf,
  resamples = penguin_folds,
  grid = 5
)

# Best Model
best_tree <- tune_res %>% select_best(metric = "accuracy")
final_wf <- finalize_workflow(penguin_wf, best_tree)
final_fit <- final_wf %>% fit(data = penguin_train)

# Test Performance
final_fit %>%
  predict(penguin_test) %>%
  bind_cols(penguin_test) %>%
  metrics(truth = species, estimate = .pred_class)
```

## Exercises

Based on the Nature paper, "An interpretable mortality prediction model for COVID-19 patients":

1.  Load the COVID-19 patient data.
2.  Fit a `RandomForestClassifier` to predict mortality.
3.  Evaluate the model using a confusion matrix and F1 score.
4.  Plot the feature importances to identify the key biomarkers for mortality.
