---
title: "Week 5 - Machine Learning Solutions (Part 1 - Continued)"
output: html_document
---

# Week 5 - Machine Learning Solutions (PART 1 - Continued)

In this session, we continue with solutions for the `diabetes` dataset, exploring both linear models and Random Forests using the `tidymodels` framework.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(lars)
library(ranger)
library(vip)
```

```{r}
# Load diabetes data
data(diabetes)
diabetes_df <- as_tibble(diabetes$x) %>%
  mutate(target = diabetes$y)
```

## Exercise 1 & 2: Multivariate Linear Regression

```{r}
set.seed(42)
diab_split <- initial_split(diabetes_df, prop = 0.7)
diab_train <- training(diab_split)
diab_test  <- testing(diab_split)

# Spec
lm_spec <- linear_reg() %>% set_engine("lm")

# Fit
lm_fit <- lm_spec %>% fit(target ~ ., data = diab_train)

# Evaluate
lm_results <- diab_test %>%
  bind_cols(predict(lm_fit, diab_test))

metrics(lm_results, truth = target, estimate = .pred)
```

## Exercise 3: Random Forest

```{r}
# Use a subset of predictors as suggested
rf_spec <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_fit <- rf_spec %>%
  fit(target ~ age + bmi + bp, data = diab_train)

# Evaluate
rf_results <- diab_test %>%
  bind_cols(predict(rf_fit, diab_test))

metrics(rf_results, truth = target, estimate = .pred)
```

## Exercise 4: Hyperparameter Tuning

We will tune `mtry` and `min_n` using a grid search.

```{r}
# Tuneable spec
rf_tune_spec <- rand_forest(
  mtry = tune(),
  trees = 100,
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Workflow
rf_wf <- workflow() %>%
  add_formula(target ~ .) %>%
  add_model(rf_tune_spec)

# Folds
set.seed(123)
diab_folds <- vfold_cv(diab_train, v = 5)

# Grid
rf_grid <- grid_regular(
  mtry(range = c(1, 10)),
  min_n(range = c(2, 10)),
  levels = 5
)

# Tuning
tune_results <- tune_grid(
  rf_wf,
  resamples = diab_folds,
  grid = rf_grid
)

# Visualization
autoplot(tune_results)
```

### Final Model Selection

```{r}
best_params <- tune_results %>% select_best(metric = "rmse")
final_rf <- finalize_workflow(rf_wf, best_params)
final_fit <- fit(final_rf, diab_train)

# Performance on test set
final_fit %>%
  predict(diab_test) %>%
  bind_cols(diab_test) %>%
  metrics(truth = target, estimate = .pred)
```
