---
title: "Week 7 - METABRIC Data Analysis"
output: html_document
---

# METABRIC - Data Analysis

**By Katarzyna Nurzynska (Adapted for R)**

The Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) database.
Obtained from the following publication: https://www.nature.com/articles/ncomms11479

### Aims

1.  **To identify factors that contribute to patient survival.**
2.  **To examine the utility of dimensionality reduction (PCA) and K-means clustering to identify subgroups of patients.**
3.  **To use Machine Learning tools (Linear Regression and Survival Analysis) to describe factors in cancer heterogeneity.**

```{r setup, message=FALSE, warning=FALSE}
# Import libraries
library(tidyverse)
library(tidymodels)
library(survival)
library(survminer)
library(broom)
library(ggpubr)
library(factoextra)
library(cluster)
library(corrplot)
```

## Data Overview

```{r}
# Input data
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv")

# Dimensions
dim(metabric)

# Summary
summary(metabric)
```

### Visualisation

```{r}
p1 <- ggplot(metabric, aes(x = Vital_status)) + geom_bar() + theme_minimal()
p2 <- ggplot(metabric, aes(x = as.factor(Tumour_stage))) + geom_bar() + labs(x = "Tumour Stage") + theme_minimal()
p3 <- ggplot(metabric, aes(x = PAM50)) + geom_bar() + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggarrange(p1, p2, p3, ncol = 2, nrow = 2)
```

## Dimensionality Reduction (PCA)

We preprocess the data by handling missing values and encoding categorical variables.

```{r}
# Remove NAs and non-numeric columns for PCA
metabric_clean <- metabric %>% drop_na()

pca_data <- metabric_clean %>%
  select(Age_at_diagnosis, Tumour_size, Tumour_stage, Mutation_count, ESR1, ERBB2, PGR, TP53, PIK3CA, GATA3, FOXA1, MLPH)

# Scale and PCA
pca_res <- prcomp(pca_data, scale. = TRUE)

# Scree plot
fviz_eig(pca_res, addlabels = TRUE)
```

### PCA Biplot

```{r}
fviz_pca_biplot(pca_res,
                label = "var",
                habillage = as.factor(metabric_clean$PAM50),
                addEllipses = TRUE,
                col.var = "red",
                alpha.ind = 0.5) +
  theme_minimal()
```

## Clustering

```{r}
# K-means clustering on first two PCs
set.seed(42)
km_res <- kmeans(pca_res$x[,1:2], centers = 4, nstart = 25)

# Visualise clusters
fviz_cluster(km_res, data = pca_res$x[,1:2], palette = "jco", ggtheme = theme_minimal())
```

## Survival Analysis

### Kaplan-Meier Curves

```{r}
# Create survival object
# Status: 0=Living, 1=Deceased (simplified for demo)
metabric_clean <- metabric_clean %>%
  mutate(status = if_else(Survival_status == "DECEASED", 1, 0))

fit <- survfit(Surv(Survival_time, status) ~ PAM50, data = metabric_clean)

ggsurvplot(fit, data = metabric_clean, pval = TRUE, title = "Survival by PAM50 Subtype")
```

### Cox Proportional Hazards Model

```{r}
cox_fit <- coxph(Surv(Survival_time, status) ~ Tumour_size + Tumour_stage + Age_at_diagnosis, data = metabric_clean)
summary(cox_fit)
```

## Summary

- PCA identified clear distinctions between LumA/B and Basal subtypes.
- Survival analysis confirmed that clinical features like tumour size and stage, along with molecular subtype, are significant predictors of outcome.
