---
title: "Week 6 - Machine Learning Solutions (PART 2)"
output: html_document
---

# Week 6 - Machine Learning Solutions (PART 2)

In this session, we implement a classification model based on the Nature Machine Intelligence paper, "An interpretable mortality prediction model for COVID-19 patients".

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(ranger)
library(vip)
```

## 1. Data Preparation

```{r}
# Note: In a real scenario, we would load the dataset here
# For this solution, we assume a data frame 'covid_data' exists
# covid_data <- read_csv("../data/covid_mortality_data.csv")

# Simulated structure for demonstration:
set.seed(42)
covid_data <- tibble(
  LDH = rnorm(200, 400, 100),
  hs_CRP = rnorm(200, 50, 20),
  lymphocytes = rnorm(200, 15, 5),
  outcome = factor(sample(c("Survived", "Deceased"), 200, replace = TRUE))
)
```

## 2. Model Training and Evaluation

```{r}
# 1. Split
covid_split <- initial_split(covid_data, prop = 0.7, strata = outcome)
covid_train <- training(covid_split)
covid_test  <- testing(covid_split)

# 2. Model Specification
rf_spec <- rand_forest(trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# 3. Fit
rf_fit <- rf_spec %>%
  fit(outcome ~ ., data = covid_train)

# 4. Predict and Evaluate
results <- covid_test %>%
  select(outcome) %>%
  bind_cols(predict(rf_fit, covid_test)) %>%
  bind_cols(predict(rf_fit, covid_test, type = "prob"))

# Confusion Matrix
conf_mat(results, truth = outcome, estimate = .pred_class)

# F1 Score
f_meas(results, truth = outcome, estimate = .pred_class)
```

## 3. Feature Importance

One of the key findings of the paper was the importance of LDH, lymphocytes, and hs-CRP.

```{r}
vip(rf_fit) +
  labs(title = "Feature Importance for COVID-19 Mortality Prediction") +
  theme_minimal()
```

## 4. Hyperparameter Tuning

```{r}
# Tuneable spec
tune_spec <- rand_forest(
  mtry = tune(),
  trees = 500,
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Folds
covid_folds <- vfold_cv(covid_train, v = 5)

# Tune
tune_res <- tune_grid(
  tune_spec,
  outcome ~ .,
  resamples = covid_folds,
  grid = 5
)

# Best Model
best_rf <- tune_res %>% select_best(metric = "roc_auc")
final_rf <- finalize_model(tune_spec, best_rf)
final_fit <- final_rf %>% fit(outcome ~ ., data = covid_train)

# Final Test Performance
final_fit %>%
  predict(covid_test) %>%
  bind_cols(covid_test) %>%
  metrics(truth = outcome, estimate = .pred_class)
```
