---
title: "Week 4 - Statistical Tests and Modelling Solutions"
output: html_document
---

# Week 4 Solutions

In this session, we provide solutions to the statistical analysis exercises using R.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(ggpubr)
```

```{r}
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv") %>%
  drop_na()
```

## Exercise 1
**Is there a significant difference between the tumour size of patients who received chemotherapy versus those who did not?**

```{r}
# Normality test
shapiro.test(metabric$Tumour_size[1:1000]) # Likely non-normal

# Visualization
ggplot(metabric, aes(x = Chemotherapy, y = Tumour_size, fill = Chemotherapy)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.2) +
  theme_minimal()

# Mann-Whitney U test (non-parametric)
wilcox.test(Tumour_size ~ Chemotherapy, data = metabric)
```

## Exercise 2
**Is there a correlation between tumour size and survival time? Does log10 transformation affect it?**

```{r}
# Correlation for raw data
cor.test(metabric$Tumour_size, metabric$Survival_time, method = "spearman")

# Correlation for log10 data
cor.test(log10(metabric$Tumour_size), metabric$Survival_time, method = "spearman")

# Visualization
ggplot(metabric, aes(x = log10(Tumour_size), y = Survival_time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()
```

## Exercise 3
**Investigate if Chemotherapy and Radiotherapy treatments are independent.**

```{r}
# Contingency table
cont_table <- table(metabric$Chemotherapy, metabric$Radiotherapy)
print(cont_table)

# Chi-square test
chisq.test(cont_table)
```
