---
title: "Week 5 - An introduction to Machine Learning (PART 1)"
output: html_document
---

# Week 5 - An introduction to Machine Learning (PART 1)

### Agenda

1. Motivation
2. Aims
3. Theoretical part
4. Work with datasets

## Motivation

> Machine Learning is a toolkit of techniques that allow us to describe data and make predictions, while making fewer assumptions about the statistical properties of our data. In R, the `tidymodels` framework provides a modern, cohesive system for machine learning that follows `tidyverse` principles.

## Aims

#### This week

- Theoretical lecture:
  - Comparing ML with statistical models
  - Types of problems where ML is helpful
  - Stages of creating an ML model
  - Introduction to the `tidymodels` framework in R
- Recreate simple and multivariate regressions based on the METABRIC dataset.
- Introduction to Decision Trees and Random Forests using the CPS Wages dataset.

## Setup

We will use the `tidymodels` suite, which includes packages like `rsample`, `parsnip`, `recipes`, and `workflows`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(ranger)
library(rpart)
library(rpart.plot)
library(vip)
```

## 1. Linear Regression with tidymodels

We will predict survival time based on tumour size using the METABRIC dataset.

### Load Data

```{r}
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv") %>%
  drop_na()
```

### Visualisation

```{r}
ggplot(metabric, aes(x = Tumour_size, y = Survival_time, color = Cancer_type)) +
  geom_point(alpha = 0.5) +
  labs(title = "Survival time vs. Tumour size", x = "Tumour size (mm)", y = "Survival time (days)") +
  theme_minimal()
```

### Modeling Workflow

In `tidymodels`, we follow a structured workflow:

1.  **Split** the data.
2.  Define the **model specification** (engine and mode).
3.  **Fit** the model.
4.  **Evaluate** predictions.

```{r}
# 1. Split data
set.seed(42)
metabric_split <- initial_split(metabric, prop = 0.8)
train_data <- training(metabric_split)
test_data  <- testing(metabric_split)

# 2. Model specification
lm_spec <- linear_reg() %>%
  set_engine("lm")

# 3. Fit model
lm_fit <- lm_spec %>%
  fit(Survival_time ~ Tumour_size, data = train_data)

# 4. Predict and Evaluate
results <- test_data %>%
  select(Survival_time, Tumour_size) %>%
  bind_cols(predict(lm_fit, test_data))

metrics(results, truth = Survival_time, estimate = .pred)
```

## 2. Decision Trees and Random Forest

We will explore non-linear models using the CPS Wages dataset.

```{r}
# Loading CPS Wages data
wages <- read_table("http://lib.stat.cmu.edu/datasets/CPS_85_Wages",
                    skip = 27, col_names = FALSE) %>%
  select(X1, X4, X6, X7) # Simplified selection for demo
colnames(wages) <- c("EDUCATION", "EXPERIENCE", "WAGE", "AGE")
```

### Decision Tree

```{r}
tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("regression")

tree_fit <- tree_spec %>%
  fit(WAGE ~ EDUCATION + AGE, data = wages)

# Visualize the tree
rpart.plot(tree_fit$fit)
```

### Random Forest

```{r}
rf_spec <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_fit <- rf_spec %>%
  fit(WAGE ~ EDUCATION + AGE, data = wages)

# Feature Importance
vip(rf_fit)
```

## Exercises

1.  Using `tidymodels`, fit a multivariate linear regression on the METABRIC dataset using `Tumour_size`, `ER_status`, and `Cancer_type` as predictors.
2.  Use `recipes` to handle the one-hot encoding of categorical variables (`step_dummy`).
3.  Compare the performance of a Linear Regression model vs. a Random Forest model on the `diabetes` dataset.
