---
title: "Week 1 - Assignment Solutions"
output: html_document
---

# Week 1 - assignment solution

### Exercise 1

- Read the dataset `metabric_clinical_and_expression_data.csv` and store its summary statistics into a new variable called `metabric_summary`.
- Write your summary statistics object into a file called `metabric_summary.csv`.
- Generate a Tab Separated Value (TSV) file instead.
- Produce an Excel spreadsheet containing summary statistics using `writexl`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(writexl)
```

```{r}
# Read metabric dataset
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv")

# Store summary statistics
# In R, summary() returns a table, but for writing to file, we might want a tidy summary
metabric_summary <- summary(metabric)
print(metabric_summary)

# Better summary for export:
metabric_stats <- metabric %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd), na.rm = TRUE))
```

```{r}
# Write summary statistics
write_csv(metabric_stats, "metabric_summary.csv")
write_tsv(metabric_stats, "metabric_summary.tsv")
write_xlsx(metabric_stats, "metabric_summary.xlsx")
```

### Exercise 2

- Calculate the mean tumour size of patients grouped by vital status and tumour stage.
- Find the cohort of patients and tumour stage where the average expression of genes TP53 and FOXA1 is highest.
- Do patients with greater tumour size live longer? How about patients with greater tumour stage? How about greater Nottingham_prognostic_index?

```{r}
# Grouped mean tumour size
metabric %>%
  group_by(Vital_status, Tumour_stage) %>%
  summarise(mean_size = mean(Tumour_size, na.rm = TRUE), .groups = "drop")
```

```{r}
# Highest average expression of TP53 and FOXA1
metabric %>%
  mutate(gene_mean = (TP53 + FOXA1) / 2) %>%
  group_by(Cohort, Tumour_stage) %>%
  summarise(avg_expr = mean(gene_mean, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(avg_expr)) %>%
  slice(1)
```

```{r}
# Correlations with Survival Time
metabric_dead <- metabric %>% filter(Vital_status == 'Died of Disease')

# Tumour size correlation
cor(metabric_dead$Tumour_size, metabric_dead$Survival_time, use = "complete.obs")

# Tumour stage grouped stats
metabric_dead %>%
  group_by(Tumour_stage) %>%
  summarise(mean_surv = mean(Survival_time, na.rm = TRUE), n = n())

# Nottingham Prognostic Index correlation
cor(metabric_dead$Nottingham_prognostic_index, metabric_dead$Survival_time, use = "complete.obs")
```

### Exercise 3 (bonus)

- Which variables have missing data?
- Find the patients IDs who have missing tumour size and/or missing mutation count data.
- Fill in missing tumour size data using the cohort average.

```{r}
# Variables with missing data
colSums(is.na(metabric))
```

```{r}
# Patients with missing values
metabric %>%
  filter(is.na(Tumour_size) | is.na(Mutation_count)) %>%
  select(Patient_ID, Cohort, Tumour_size, Mutation_count)
```

```{r}
# Fill missing tumour size with cohort mean
metabric_filled <- metabric %>%
  group_by(Cohort) %>%
  mutate(Tumour_size = if_else(is.na(Tumour_size), mean(Tumour_size, na.rm = TRUE), Tumour_size)) %>%
  ungroup()

# Validation
sum(is.na(metabric_filled$Tumour_size))
```
