---
title: "Week 5 - Machine Learning (PART 1 - Continued)"
output: html_document
---

# Week 5 - Machine Learning (PART 1 - Continued)

In this session, we continue our exploration of machine learning with a focus on non-linear models and multivariate regression using the `tidymodels` framework.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(ranger)
library(rpart)
library(rpart.plot)
library(vip)
```

## 1. Multivariate Linear Regression with Recipes

We will use the METABRIC dataset and build a preprocessing recipe to handle categorical variables.

```{r}
# Load data
metabric <- read_csv("../data/metabric_clinical_and_expression_data.csv") %>%
  drop_na()

# Data split
set.seed(42)
metabric_split <- initial_split(metabric, prop = 0.8)
train_data <- training(metabric_split)
test_data  <- testing(metabric_split)

# Recipe: Preprocessing
metabric_recipe <- recipe(Survival_time ~ Tumour_size + ER_status + Cancer_type, data = train_data) %>%
  step_dummy(ER_status, Cancer_type) # One-hot encoding equivalent

# Model spec
lm_spec <- linear_reg() %>% set_engine("lm")

# Workflow
metabric_wflow <- workflow() %>%
  add_recipe(metabric_recipe) %>%
  add_model(lm_spec)

# Fit
metabric_fit <- fit(metabric_wflow, train_data)

# Results
tidy(metabric_fit)
```

## 2. Moving Beyond Linear Models

We explore Decision Trees and Random Forests using the CPS Wages dataset.

```{r}
# Load CPS Wages
wages_url <- "http://lib.stat.cmu.edu/datasets/CPS_85_Wages"
wages <- read_table(wages_url, skip = 27, col_names = FALSE) %>%
  select(X1, X4, X6, X7)
colnames(wages) <- c("EDUCATION", "EXPERIENCE", "WAGE", "AGE")
```

### Random Forest Regression

```{r}
# Model spec
rf_spec <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# Fit
rf_fit <- rf_spec %>%
  fit(WAGE ~ EXPERIENCE + EDUCATION + AGE, data = wages)

# Feature Importance
vip(rf_fit)
```

## Exercises

1.  Compare the performance (RMSE) of a Linear Regression model versus a Random Forest model on the `wages` dataset.
2.  Use `rsample` to perform 5-fold cross-validation on your Random Forest model.
3.  Tune the `trees` hyperparameter for the Random Forest model.
