---
title: "Week 7 - Networks and Knowledge Graphs"
output: html_document
---

# R Course - Networks and Knowledge Graphs

Katarzyna Nurzynska, Sebastian Nilsson, BIKG (Adapted for R)

## Overview

**We will go through:**

*   Background on networks and graphs
*   **Part 1: Simple Networks**
    *   Building networks with `tidygraph`
    *   Network analysis using degree metrics
    *   Visualization with `ggraph` and `visNetwork`
    *   *First use case: drug-drug interactions*
*   **Part 2: Knowledge Graphs & BIKG**
    *   Introduction to Knowledge Graphs
    *   Overview of the Biological Insights Knowledge Graph (BIKG)
    *   *Second use case: Trop2*

## Background

**What are networks and graphs?**

A network (or graph) is a collection of nodes (vertices) and edges (links) that model relationships. In biological systems, these can represent protein-protein interactions, metabolic pathways, or drug-drug interactions.

**What is a Knowledge Graph?**

A Knowledge Graph is a set of data points connected by relations that describe a domain. It enables automated insights and complex queries across heterogeneous data sources.

## Part 1: Simple Networks in R

In R, the `tidygraph` package provides a tidy approach to graph manipulation, while `ggraph` extends `ggplot2` for graph visualization.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidygraph)
library(ggraph)
library(visNetwork)
```

### Building a Graph

```{r}
# Define nodes and edges
nodes <- tibble(name = c('A', 'B', 'C', 'D', 'E'))
edges <- tibble(
  from = c('A', 'A', 'B', 'B', 'B', 'C'),
  to   = c('B', 'C', 'D', 'E', 'A', 'E')
)

# Create a tidygraph object
G <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE)

# Print vertex and edge sets
G
```

### Visualising the Graph

```{r}
ggraph(G, layout = "stress") +
  geom_edge_link() +
  geom_node_point(size = 8, color = "lightblue") +
  geom_node_text(aes(label = name), vjust = 0.5, hjust = 0.5) +
  theme_graph()
```

### Weighted Graphs

```{r}
edges_weighted <- tibble(
  from = c('A', 'A', 'B', 'B', 'C'),
  to   = c('B', 'C', 'D', 'E', 'E'),
  weight = c(2, 1, 5, 3, 2)
)

G_weighted <- tbl_graph(nodes = nodes, edges = edges_weighted, directed = FALSE)

ggraph(G_weighted, layout = "stress") +
  geom_edge_link(aes(width = weight, label = weight),
                 angle_calc = 'along', label_dodge = unit(2.5, 'mm')) +
  geom_node_point(size = 5) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_graph()
```

## First Use Case: Drug-Drug Interactions (DDI)

### 1) Data Overview

We use data summarising drug-drug interactions (DDI). Each interaction is a relationship between two compounds.

```{r, eval=FALSE}
# Example loading
# DDI_data <- read_csv("../data/DDI_data.csv")
```

### 2) Network Analysis Logic

In R, we can calculate node degree (number of connections) to identify highly interacting drugs.

```{r, eval=FALSE}
# Top 20 connected drugs example
G_ddi %>%
  activate(nodes) %>%
  mutate(degree = centrality_degree()) %>%
  arrange(desc(degree)) %>%
  head(20)
```

### 3) Interactive Visualisation with `visNetwork`

`visNetwork` allows for interactive exploration of complex networks.

```{r}
# Interactive example
nodes_vis <- data.frame(id = 1:5, label = LETTERS[1:5])
edges_vis <- data.frame(from = c(1, 2, 2, 3), to = c(2, 3, 4, 1))

visNetwork(nodes_vis, edges_vis) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visPhysics(solver = "forceAtlas2Based")
```

## Part 2: Knowledge Graphs & BIKG

Knowledge Graphs like BIKG (Biological Insights Knowledge Graph) serve as a backbone for data integration, enabling discovery via machine learning.

### Trop2 Subgraph Analysis

Trop2 is a key target in many cancer types. By querying a Knowledge Graph, we can find associated compounds and diseases.

```{r, eval=FALSE}
# Porting logic: filtering subgraphs in R
# trop2_neighbors <- edges %>%
#   filter(source_id == "TACSTD2" | target_id == "TACSTD2")
```

### Summary

*   Graphs allow us to model complex biological relationships.
*   `tidygraph` and `ggraph` bring the power of the `tidyverse` to network science.
*   Knowledge Graphs provide a platform for data integration and emergent signal detection.
