---
title: "Week 5 - Machine Learning Solutions (Part 1)"
output: html_document
---

# Week 5 - Machine Learning Solutions (PART 1)

In this session, we work with the `diabetes` dataset to predict disease progression. In R, we use the `tidymodels` framework.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(lars) # For the diabetes dataset
```

```{r}
# Load diabetes data
data(diabetes)
diabetes_df <- as_tibble(diabetes$x) %>%
  mutate(target = diabetes$y)

glimpse(diabetes_df)
```

## Exercise 1

Fit a multivariate linear regression using all samples to predict the response variable.

```{r}
# Model specification
lm_spec <- linear_reg() %>%
  set_engine("lm")

# Fit model
lm_fit <- lm_spec %>%
  fit(target ~ ., data = diabetes_df)

# Results
tidy(lm_fit)
glance(lm_fit)
```

## Exercise 2

Perform a test-train split and evaluate the model.

```{r}
set.seed(42)
diab_split <- initial_split(diabetes_df, prop = 0.7)
diab_train <- training(diab_split)
diab_test  <- testing(diab_split)

# Fit on training set
lm_fit_split <- lm_spec %>%
  fit(target ~ ., data = diab_train)

# Predict on test set
results_test <- diab_test %>%
  bind_cols(predict(lm_fit_split, diab_test))

# Metrics
metrics(results_test, truth = target, estimate = .pred)
```

## Exercise 3

Try using a Random Forest model instead.

```{r}
library(ranger)

rf_spec <- rand_forest(trees = 100) %>%
  set_engine("ranger") %>%
  set_mode("regression")

rf_fit <- rf_spec %>%
  fit(target ~ ., data = diab_train)

# Evaluate
results_rf <- diab_test %>%
  bind_cols(predict(rf_fit, diab_test))

metrics(results_rf, truth = target, estimate = .pred)
```

## Exercise 4: Hyperparameter Tuning

Tweaking `mtry` and `min_n` using `tidymodels` grid search.

```{r}
# Tuneable specification
rf_tune_spec <- rand_forest(
  mtry = tune(),
  trees = 100,
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Cross-validation folds
set.seed(123)
diab_folds <- vfold_cv(diab_train, v = 5)

# Tuning grid
rf_grid <- grid_regular(
  mtry(range = c(1, 10)),
  min_n(range = c(2, 10)),
  levels = 5
)

# Run tuning
tune_results <- tune_grid(
  rf_tune_spec,
  target ~ .,
  resamples = diab_folds,
  grid = rf_grid
)

autoplot(tune_results)
```

```{r}
# Best model
best_rf <- tune_results %>%
  select_best(metric = "rmse")

final_rf <- finalize_model(rf_tune_spec, best_rf)
final_fit <- final_rf %>%
  fit(target ~ ., data = diab_train)

# Final performance
final_results <- diab_test %>%
  bind_cols(predict(final_fit, diab_test))

metrics(final_results, truth = target, estimate = .pred)
```
